Parsing Postfix log files

### 2009-04-08 doc/logparser-complications.tex
Final corrections/improvements based on Carl's comments.

Each of these complications caused \parsername{} to operate incorrectly,
generate warning messages, or leave mails in the state table.  The
complications are listed in the order in which they were overcome during
development of \parsername{}, with the first complication occurring several
orders of magnitude more frequently than the last.  When deciding which
problem to tackle next, the problem causing the highest number of warning
messages or mails incorrectly remaining in the state tables was always
chosen, because that approach yielded the biggest improvement in the
parser, and made the remaining problems more apparent.

>>>>

Each of these complications caused \parsername{} to operate incorrectly,
generate warning messages, or leave mails in the state table.  The
complications are listed in the order in which they were overcome during
development of \parsername{}, with the first complication occurring several
orders of magnitude more frequently than the last.  When deciding which
problem to tackle next, the problem causing the highest number of warning
messages or mails incorrectly remaining in the state tables was always
chosen, because that approach yielded the biggest improvement in the
parser, and made the remaining problems more apparent.  Several of the
complications were discovered because \parsername{} is very careful to
check the origin of each mail; if it does not know the origin of a mail,
that mail will be tagged as fake, and a warning issued if the mail's origin
has not been determined before trying to save it to the database.  Some of
the solutions to these complications require discarding log lines because
they are out of order, and, less frequently, discarding a data structure
containing data gathered about a connection or a mail, because of the
paucity of information contained in the data structure.

### 2009-04-08 doc/logparser-implementation.tex
Final corrections/improvements based on Carl's comments.

\parsername{} deals with all the complications of parsing, the
eccentricities and oddities of Postfix log files, presenting the resulting
data in a normalised, simple to use representation.  The parser's task is
to follow the journey each mail takes through Postfix, combining the data
captured by rules into a coherent whole, saving it in a useful and
consistent form, and performing housekeeping duties.  The intermingling of
log lines from different mails immediately rules out the possibility of
handling each mail in isolation; the parser must be able to handle multiple
mails in parallel, each potentially at a different stage in its journey,
without any interference between mails --- except in the minority of cases
where intra-mail interference is required, e.g.\ mail re-injected for
forwarding (\sectionref{Re-injected mails}).  The best way to deal with
intermingling is to maintain state information for every unfinished mail
and manipulate the appropriate mail correctly for each log line
encountered.

@@

The flow chart in \figureref{flow chart image} shows the most common paths
a connection or mail can take through \parsername{}; the difficulties
described in \sectionref{complications} are excluded for the sake of
clarity.  The flow chart is intended to be a graphical overview of how a
mail progresses through both Postfix and \parsername{}, providing an
overall context into which the detailed descriptions in the remainder of
this chapter will fit, in particular the actions (\sectionref{actions in
implementation}) and complications (\sectionref{complications}).  The
states and transitions shown in \figureref{flow chart image} will be
explained later in this section.

@@

Each time the parser is run the framework performs some initialisation
tasks, setting up data structures that will later be used by actions,
including several state tables.  Data about the connections and mails being
processed is held in \texttt{connections} and \texttt{queueids}
respectively.  The other state tables used when working around
complications in parsing: \texttt{timeout\_queueids} is used when dealing
with connections that time out during the DATA phase (\sectionref{timeouts
during data phase}); \texttt{bounce\_queueids} is part of the solution to
bounce notification mails being delivered before their creation is logged
(\sectionref{Bounce notification mails delivered before their creation is
logged}); and \texttt{postsuper\_deleted\_queueids} caches information
about mails recently deleted by the administrator so that subsequent log
lines processed by the \action{SAVE\_DATA} action can be discarded
(\sectionref{Mail deleted before delivery is attempted}).

@@

The framework is now ready to begin parsing.  For each log line it will use
those rules whose \texttt{program} field equals the program in the log line
(previously described in \sectionref{rule conditions in implementation}),
falling back to generic rules if necessary, and warning if the log line is
unrecognised.  The rules are sorted for increased efficiency; see
\sectionref{rule ordering for efficiency} for details.  The repetitive
nature of log files gives them high compression ratios; the framework uses
a Perl module named \texttt{IO::Uncompress::AnyUncompress} to read
compressed log files; this saves users the trouble of uncompressing them to
a temporary file before parsing begins.  When used interactively, the
framework also displays a progress bar to show how far parsing has
progressed through the log file and how long the remainder of the parsing
process is likely to take.  The progress bar is not as reliable when
parsing compressed log files, because a compressed block may uncompress to
a variable number of log lines; there is also variation between the
recognition and processing time of individual log lines, but overall the
progress bar is a useful addition.

The framework collects data used when evaluating \parsernames{} efficiency
for the results chapter (\sectionref{Results}); some of the techniques used
to improve parsing speed can be turned off or altered to measure the effect
they have.  Three sets of data are collected:

XXX ASK CARL IF I SHOULD HAVE MEAN, STANDARD DEVIATION, MORE GRAPHS\@?

@@

There are five ways the framework can adapt its behaviour to demonstrate
how effective the optimisations it uses are, and how efficient
\parsername{} is:

@@

    \item The framework can skip inserting data into the database, because
        that dominates the execution time of the parser; the evaluation
        chapter measures the speed of \parsername{}, not the speed of the
        database or the disks it resides on.

@@

This utility is not intended to create perfect regexes, but it greatly
reduces the effort required to deal with unrecognised log lines.

@@

>>>>

\parsername{} deals with all the complications of parsing, the
eccentricities and oddities of Postfix log files, presenting the resulting
data in a normalised, simple to use representation.  Unfortunately dealing
with the complications that arise sometimes requires the parser to discard
log lines (e.g.\ \sectionref{aborted delivery attempts}), and, less
frequently, to discard a data structure (e.g.\ \sectionref{timeouts during
data phase}).  \parsername{} is capable of parsing log files from Postfix
version 2.2 through to version 2.5, and should be capable of parsing log
files from later versions with only minor modifications or an updated
ruleset.  The parser's task is to follow the journey each mail takes
through Postfix, combining the data captured by rules into a coherent
whole, saving it in a useful and consistent form, and performing
housekeeping duties.  The intermingling of log lines from different mails
immediately rules out the possibility of handling each mail in isolation;
the parser must be able to handle multiple mails in parallel, each
potentially at a different stage in its journey, without any interference
between mails --- except in the minority of cases where intra-mail
interference is required, e.g.\ mail re-injected for forwarding
(\sectionref{Re-injected mails}).  The best way to deal with intermingling
of log lines is to maintain state information for every unfinished mail,
and to manipulate the appropriate mail correctly for each log line
encountered.

@@

The flow chart in \figureref{flow chart image} shows the most common paths
a connection or mail can take through \parsername{}; decision boxes and the
difficulties described in \sectionref{complications} are excluded for the
sake of clarity.  The flow chart is intended to be a graphical overview of
how a mail progresses through both Postfix and \parsername{}, providing an
overall context into which the detailed descriptions in the remainder of
this chapter will fit, in particular the actions (\sectionref{actions in
implementation}) and complications (\sectionref{complications}).  The
states and transitions shown in \figureref{flow chart image} will be
explained later in this section.

@@

Each time the parser is run the framework performs some initialisation
tasks, setting up data structures that will later be used by actions,
including several state tables.  Data about the connections and mails being
processed is held in \texttt{connections} and \texttt{queueids}
respectively.  The other state tables are used when working around
complications in parsing: \texttt{timeout\_queueids} is used when dealing
with connections that time out during the DATA phase (\sectionref{timeouts
during data phase}); \texttt{bounce\_queueids} is part of the solution to
bounce notification mails being delivered before their creation is logged
(\sectionref{Bounce notification mails delivered before their creation is
logged}); and \texttt{postsuper\_deleted\_queueids} caches information
about mails recently deleted by the administrator so that subsequent log
lines processed by the \action{SAVE\_DATA} action can be discarded
(\sectionref{Mail deleted before delivery is attempted}).

@@

The framework is now ready to begin parsing.  For each log line it will use
those rules whose \texttt{program} field equals the program in the log line
(previously described in \sectionref{rule conditions in implementation}),
falling back to generic rules if necessary, and warning if the log line is
unrecognised.  The rules are sorted for increased efficiency; see
\sectionref{rule ordering for efficiency} for details.  The repetitive
nature of log files gives them high compression ratios; the framework uses
a Perl module named \texttt{IO::Uncompress::AnyUncompress} to read
compressed log files; this saves users the trouble of uncompressing them to
a temporary file before parsing begins.  When used interactively, the
framework also displays a progress bar to show how far parsing has
progressed through the log file and how long the remainder of the parsing
process is likely to take.  The progress bar is not as reliable when
parsing compressed log files, because a compressed block may uncompress to
a variable number of log lines; there is also variation between the
recognition and processing time of individual log lines, but overall the
progress bar is a useful addition.  \parsername{} is primarily intended for
parsing complete log files, but with minor modifications it could be used
to parse a live log file, periodically checking it for new log lines; this
could be very useful for programs that work best with up to date data,
e.g.\ a program for monitoring the health of the mail system.

The framework collects data used when evaluating \parsernames{} efficiency
for the results chapter (\sectionref{Results}); some of the techniques used
to improve parsing speed can be turned off or altered to measure the effect
they have.  In each case the data gathered with the optimisation disabled
will be compared to the data gathered with it enabled, to quantify the
benefit each optimisation provides.  Three sets of data are collected:

@@

There are five ways the framework can adapt its behaviour to demonstrate
how effective each optimisation is:

@@

    \item The framework can skip inserting data into the database, because
        that dominates the execution time of the parser; the evaluation
        chapter measures the speed of \parsername{}, not the speed of the
        database or the disks it resides on.  The impact of storing results
        in the database is described at the beginning of
        \sectionref{Results}.

@@

This utility is not intended to create perfect regexes, but it greatly
reduces the effort required to deal with unrecognised log lines.  The
regexes it generates will be self-contained: a parser that relies on using
cascaded parsing would require a modified algorithm, perhaps by replacing
the pre-processing stage with one that applies existing cascading rules to
each log line, and uses the resulting modified log lines for the remainder
of the algorithm.

@@

There is some similarity between regex components and cascaded parsing:
each regex component resembles a rule that recognises part of a log line
and consumes it, leaving the remainder to be recognised by other components
or cascaded rules.  The major difference between the two is that regex
components are explicitly used by the author of the ruleset, whereas
cascaded parsing would be dynamically applied by the framework.
